---
title: "Naive Bayes Classifier"
author: "Prajwal C N"
date: "04/03/2021"
output: 
  html_document:
    theme: readable
    highlight: haddock
---

#### Loading R libraries
```{r, warning=FALSE, message=FALSE}
library(readr)
library(reshape2)
library(dplyr)
library(e1071)
library(ggplot2)
library(caret)
library(class)
library(gmodels)
library(pROC)
```


#### Importing the UniversalBank dataset
```{r,message=FALSE}
UniversalBank <- read_csv("UniversalBank.csv")

```


Given we focus on two predictors: Online and Credit Card, and the outcome Personal Loan. So Converting these variables to factors
```{r,message=FALSE}
UniversalBank <- UniversalBank %>% mutate(`Personal Loan` = as.factor(`Personal Loan`), 
                         Online = as.factor(Online),
                         CreditCard = as.factor(CreditCard))

#Examining the dataset
glimpse(UniversalBank)
```


#### Data cleaning and Exploratory Data Analysis
Checking summary for each categorical variable
```{r}
summary(UniversalBank %>%  select(`Personal Loan`, Online, CreditCard))
```


Checking NULL values in the dataset at column level.
```{r}
apply(UniversalBank,2,function(x){any(is.na(x))})
```


**Visualization**


Visualizing distribution of each categorical variable.


1) Credit Card category
```{r, message=FALSE, warning=FALSE}
ggplot(UniversalBank, aes(x=CreditCard, y=..prop..,group = 1)) + 
  geom_bar(fill="light blue") +
  theme_classic() + 
  geom_text(aes(label=round(..prop..,2)),stat = "count",
            position = position_stack(vjust=0.5)) + 
  labs(y = 'Proportion', title = "Proportion of Credit Card users") +
  scale_x_discrete(labels = c("Non Card Holders","Card Holders"))
```


From the above graph we can infer that majority of population don't hold credit card, only 29% have it. 


2) Personal Loan category
```{r, message=FALSE, warning=FALSE}
ggplot(UniversalBank, aes(x=`Personal Loan`, y=..prop..,group = 1)) + 
  geom_bar(fill="cornsilk2") +
  theme_classic() + 
  geom_text(aes(label=round(..prop..,2)),stat = "count",
            position = position_stack(vjust=0.5)) + 
  labs(y = 'Proportion', title = "Proportion of Personal Loan category") + 
  scale_x_discrete(labels = c("Loan Deined","Loan Accepted"))
```


From the above graph we can see that only around 10% of population accepted loan and rest rejected it. 


3) Online category
```{r, message=FALSE, warning=FALSE}
ggplot(UniversalBank, aes(x=Online, y=..prop..,group = 1)) + 
  geom_bar(fill="azure2") + 
  theme_classic() + 
  geom_text(aes(label=round(..prop..,2)),stat = "count",
            position = position_stack(vjust=0.5)) + 
  labs(x="Online Banking users",y = 'Proportion',title = "Proportion of Online Banking users") + 
  scale_x_discrete(labels = c("Non Users","Users"))
```


From the above graph we can infer that around 60% of population using online banking services, and remaining 40% not opted for it. 


Proportion of Credit Card users
```{r}
as.data.frame(prop.table(table(UniversalBank[c("CreditCard","Personal Loan")]))) %>% 
  ggplot(aes(x=CreditCard,y=Freq,fill=Personal.Loan)) + geom_col() + 
  geom_text(aes(label=round(Freq,2)),position = position_stack(vjust = 0.5),size=2.8) + 
  theme_classic() + labs( y = 'Proportion', title = "Proportion of Credit Card users") + 
  scale_x_discrete(labels = c("Non Card Holders","Card Holders")) + 
  theme(legend.title = element_blank()) + 
  scale_fill_manual(labels = c("Loan Rejected","Loan Accepted"), 
                    values = c("azure2","light blue"))
```


Proportion of Online Banking users
```{r}
as.data.frame(prop.table(table(UniversalBank[c("Online","Personal Loan")]))) %>% 
  ggplot(aes(x=Online,y=Freq,fill=Personal.Loan)) + geom_col() + 
  geom_text(aes(label=round(Freq,2)),position = position_stack(vjust = 0.5)) + theme_classic() + 
  labs(x="Online Banking users",y = 'Proportion',title = "Proportion of Online Banking users") + 
  scale_x_discrete(labels = c("Non Users","Users")) + 
  theme(legend.title = element_blank()) + 
  scale_fill_manual(labels = c("Loan Rejected","Loan Accepted"), 
                    values = c("azure2","light blue"))
```


#### Pre-Processing of data
**Splitting dataset into training (60%) and validation (40%) sets**
```{r, message=FALSE, warning=FALSE}
set.seed(22)
index <- createDataPartition(UniversalBank$`Personal Loan`, p=0.6, list = FALSE)
UniversalBank_train_df <- UniversalBank[index,]
UniversalBank_val_df <- UniversalBank[-index,]
```


*Q1:* Creating pivot table for the training data using melt() and cast() functions.
```{r, message=FALSE, warning=FALSE}
UniversalBank_melt <- melt(UniversalBank_train_df, id = c("CreditCard","Personal Loan"), 
  variable = "Online")
UniversalBank_cast <- dcast(UniversalBank_melt, CreditCard+`Personal Loan` ~ Online)
(pivottable1 <- UniversalBank_cast %>% select(CreditCard,`Personal Loan`,Online))
```


*Q2:* Probability of Loan acceptance rate for customer who has credit card and is actively using online banking services.
```{r, message=FALSE, warning=FALSE}
(loanacceptance_rate <- (pivottable1[4,"Online"]/sum(pivottable1[4,"Online"],
                                                     pivottable1[3,"Online"])))
```


*Q3:* Creating two separate pivot tables for the training data.


*i:* Table 1 will have Loan (rows) as a function of Online (columns).
```{r, message=FALSE, warning=FALSE}
UniversalBank_melt_t1 <- melt(UniversalBank_train_df,id = c("Personal Loan"),variable = "Online")
UniversalBank_cast_t1 <- dcast(UniversalBank_melt_t1, `Personal Loan` ~ Online)
(UniversalBank_cast_t1 %>% select(`Personal Loan`,Online))
```


*ii:* Table 2 will have Loan (rows) as a function of CC.
```{r, message=FALSE, warning=FALSE}
UniversalBank_melt_t2 <- melt(UniversalBank_train_df,id = c("CreditCard"),
  variable = "Online")
UniversalBank_cast_t2 <- dcast(UniversalBank_melt_t2, CreditCard ~ Online)
UniversalBank_cast_t2 %>% select(CreditCard,Online)
```


*Q4:* Computing quantities.

*Sub Table 1:* with Credit Card(rows) and Personal Loan(columns) details.    
```{r, message=FALSE, warning=FALSE}
(st1 <- table(UniversalBank_train_df[c("CreditCard","Personal Loan")]))
```


*Sub Table 2:* with Online(rows) and Personal Loan(columns) details.    
```{r, message=FALSE, warning=FALSE}
(st2 <- table(UniversalBank_train_df[c("Online","Personal Loan")]))
```


*Sub Table 3:* only with Personal Loan count.    
```{r, message=FALSE, warning=FALSE}
(st3 <- table(UniversalBank_train_df["Personal Loan"]))
```  


*i:* P(CC = 1 | Loan = 1)  
```{r, message=FALSE, warning=FALSE}
(q1_prob <- (st1[2,2]/(st1[2,2] + st1[1,2])) * 100)
```
[84 / (84 + 204) = 29.16%]


*ii:* P(Online = 1 | Loan = 1) 
```{r, message=FALSE, warning=FALSE}
(q2_prob <- (st2[2,2]/(st2[2,2] + st2[1,2])) * 100)
```
[165 / (165 + 123) = 57.29%]


*iii:* P(Loan = 1)
```{r, message=FALSE, warning=FALSE}
(q3_prob <- (st3[2]/(st3[2] + st3[1])) * 100)
```
[288 / (288 + 2712) = 9.6%]


*iv:* P(CC = 1 | Loan = 0)
```{r, message=FALSE, warning=FALSE}
(q4_prob <- (st1[2,1]/(st1[2,1] + st1[1,1])) * 100)
```
[792 / (792 + 1920) = 29.20%]


*v:* P(Online = 1 | Loan = 0)
```{r, message=FALSE, warning=FALSE}
(q5_prob <- (st2[2,1]/(st2[2,1] + st2[1,1])) * 100)
```
[1638 / (1638 + 1074) = 60.39%]


*vi:* P(Loan = 0)
```{r, message=FALSE, warning=FALSE}
(q6_prob <- (st3[1]/(st3[1] + st3[2])) * 100)
```
[2712 / (2712 + 288) = 90.4%]


*Q5:* Using naive Bayes probability to compute P(Loan = 1 | CC = 1,Online = 1).
```{r}
(nb_prob <- (q1_prob * q2_prob * q3_prob) / ((q1_prob * q2_prob * q3_prob) + 
 (q4_prob * q5_prob * q6_prob)))
```
[((84 / (84 + 204))*(165 / (165 + 123))*(288 / (288 + 2712)))/((84 / (84 + 204))*(165 / (165 + 123))*(288 / (288 + 2712)) + (792 / (792 + 1920))*(1638 / (1638 + 1074))*(2712 / (2712 + 288))) = 9.14%]


*Q6:* Comparing above value with the one obtained from the pivot table in (B). 

The value obtained from above question(naive Bayes probability) is **0.0914** which is almost similar to pivot table(B) value **0.09589**, but pivot table(B) provides exact probability value while naive bayes has slight change in probability value, but the order will be maintained.



*Q7:* Using naive Bayes model for computing P(Loan = 1 | CC = 1, Online = 1)  
```{r}
# Build a naive Bayes classifier
nb_train <-  UniversalBank_train_df
nb_test <- UniversalBank_val_df
nb_model <-  naiveBayes(`Personal Loan`~.,data=nb_train)
```


**priori probabilities of Personal Loan** 
```{r}
prop.table(nb_model$apriori)
```


**Conditional probabilities of CreditCard** 
```{r}
nb_model$tables$CreditCard
```


**Conditional probabilities of Online**
```{r}
nb_model$tables$Online
```


From the above results we can see that priori probabilities of naive Bayes wrt loan acceptance(1) is **0.096** which is almost similar to the question 5.


Now, using the nb_model on the test set
```{r}
predictions_prob <- predict(nb_model, nb_test)
```


**Confusion Matrix of nb_model**
```{r}
confusionMatrix(nb_test$`Personal Loan`, predictions_prob)
```


**ROC Curve**
```{r}
predictions_prob1 <- predict(nb_model, nb_test, type = "raw")
roc(nb_test$`Personal Loan`, predictions_prob1[,2])
plot.roc(nb_test$`Personal Loan`, predictions_prob1[,2])
```


